Titeln på avhandlingens språk

### 2023-02-03 body/part1/background.tex
fixing formatting

\subsection{\glsfmtlong{WCA}}
\glsreset{WCA}

@@

\subsubsection{Characterization of \glsfmtshort{WCA}}

@@

\glsreset{TTF}%
From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}, label={(\arabic*)}]
    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
\end{enumerate*}
These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
We will refer to them extensively in \cref{chap:contributions}.

@@

%\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]

@@

%\end{enumerate*}

@@

\subsection{\glsfmtshortpl{NCS} and other \glsfmtshortpl{CPS}}

@@

One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
On the other, there exist several approaches when considering experimental methodologies.
%A number of works concerning \glspl{NCS} deal with experimental studies.
%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
One such approach uses setups in which the complete system is built on top of real hardware.
This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
\end{enumerate*}

@@

Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item \gls{COSMOS}
    \item \gls{POWDER}
    \item \gls{ARA}
    \item Drexel Grid \gls{SDR}
\end{enumerate*} testbeds.
\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.

>>>>

\subsection{\acl{WCA}}
\glsreset{WCA}

@@

\subsubsection{Characterization of \acs{WCA}}

@@

\glsreset{TTF}%
From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
\begin{inlineenum}
    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
\end{inlineenum}
These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
We will refer to them extensively in \cref{chap:contributions}.

@@

%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]

@@

%\end{inlineenum}

@@

\subsection{\acsp{NCS} and other \acsp{CPS}}

@@

One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
On the other, there exist several approaches when considering experimental methodologies.
%A number of works concerning \glspl{NCS} deal with experimental studies.
%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
One such approach uses setups in which the complete system is built on top of real hardware.
This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
\end{inlineenum}

@@

Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the
\begin{inlineenum}
    \item \gls{COSMOS}
    \item \gls{POWDER}
    \item \gls{ARA}
    \item Drexel Grid \gls{SDR}
\end{inlineenum} testbeds.
\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.

### 2023-02-03 publications/2019EdgeDroid/paper_body.tex
restructuring
fixing formatting

\cref{paper:olguinmunoz2019edgedroid:tbl:cloudletclienthardware} shows the hardware and software specifications for the cloudlet and the clients.
The components communicated through a single consumer-grade WiFi (IEEE 802.11n, \SI{2.4}{\giga\hertz}) access point.
We considered exclusively off-the-shelves hardware and software.
To minimize interference during the measurements, we locked the wireless link to the least congested available channel.
%Intel\textregistered \@ TurboBoost\texttrademark \@ technology was activated on the cloudlet CPU and the frequency of all cores locked to \SI{3.7}{\giga\hertz} to ensure consistency across experiments.

@@

Our measurements were obtained from series of scenarios, repeated 100 times each.
We performed three \emph{optimal} scenarios with 1, 5 and 10 clients each, the signal strength of the WiFi link being an excellent \SI{-40}{\dBm}.
Next, a scenario where the 10 clients were moved to another room roughly \SI{10}{\meter} away, thus degrading the network link to an average measured strength of \SI{-73}{\dBm}.
Finally, we studied an additional optimal single-client scenario focused on latency distributions and jitters within application execution path.
This scenario is used to showcase the utility of EdgeDroid 1.0 for application developers.

@@

We can identify a pair of interesting effects in the scaling of the task-guidance \gls{WCA} application.
One, scaling behavior for the application seems to be linear with respect to the number of clients.
Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.

>>>>

\cref{paper:olguinmunoz2019edgedroid:tbl:cloudletclienthardware} shows the hardware and software specifications for the cloudlet and the clients.
The components communicated through a single consumer-grade Wi-Fi (IEEE 802.11n, \SI{2.4}{\giga\hertz}) access point.
We considered exclusively off-the-shelves hardware and software.
To minimize interference during the measurements, we locked the wireless link to the least congested available channel.
%Intel\textregistered \@ TurboBoost\texttrademark \@ technology was activated on the cloudlet CPU and the frequency of all cores locked to \SI{3.7}{\giga\hertz} to ensure consistency across experiments.

@@

Our measurements were obtained from series of scenarios, repeated 100 times each.
We performed three \emph{optimal} scenarios with 1, 5 and 10 clients each, the signal strength of the Wi-Fi link being an excellent \SI{-40}{\dBm}.
Next, a scenario where the 10 clients were moved to another room roughly \SI{10}{\meter} away, thus degrading the network link to an average measured strength of \SI{-73}{\dBm}.
Finally, we studied an additional optimal single-client scenario focused on latency distributions and jitters within application execution path.
This scenario is used to showcase the utility of EdgeDroid 1.0 for application developers.

@@

We can identify a pair of interesting effects in the scaling of the task-guidance \gls{WCA} application.
One, scaling behavior for the application seems to be linear with respect to the number of clients.
Two, in the case of the impaired Wi-Fi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.

### 2023-02-03 publications/2021ImpactDelayedResponse/Manuscript.tex
fixing formatting

Accordingly, in the following we review background work in the areas of 
\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item[(\cref{ssec:timeperception})] time perception
  \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
\end{enumerate*}

@@

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
    \item arousal can expand durations~\cite{droit2011emotion}.
\end{enumerate*}

@@

The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
In particular
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
    \item it is easily understood by users and requires essentially no training
    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
    \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
\end{enumerate*}

@@

\subsubsection{Biometric Data}
The participants wore devices to acquire four physiological measures:
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item galvanic skin response (\emph{GSR})
  \item accelerometer data from the dominant wrist
  \item brain activity in the form of electroencephalography (\emph{EEG})
  \item heart rate.
\end{enumerate*}

@@

For the EEG data we employed the OpenBCI EEG Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
It provides a quick and non-invasive way of obtaining EEG signals from participants.
Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
The signals were sampled at \SI{200}{\hertz} and postprocessed to 
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
  \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
\end{enumerate*}

@@

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{enumerate*}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

>>>>

Accordingly, in the following we review background work in the areas of 
\begin{inlineenum}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item[(\cref{ssec:timeperception})] time perception
  \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
\end{inlineenum}

@@

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
    \item arousal can expand durations~\cite{droit2011emotion}.
\end{inlineenum}

@@

The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
In particular
\begin{inlineenum}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
    \item it is easily understood by users and requires essentially no training
    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
    \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
\end{inlineenum}

@@

\subsubsection{Biometric Data}
The participants wore devices to acquire four physiological measures:
\begin{inlineenum}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item galvanic skin response (\emph{GSR})
  \item accelerometer data from the dominant wrist
  \item brain activity in the form of electroencephalography (\emph{EEG})
  \item heart rate.
\end{inlineenum}

@@

For the EEG data we employed the OpenBCI EEG Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
It provides a quick and non-invasive way of obtaining EEG signals from participants.
Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
The signals were sampled at \SI{200}{\hertz} and postprocessed to 
\begin{inlineenum}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
  \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
\end{inlineenum}

@@

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{inlineenum}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{inlineenum}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

### 2023-02-03 publications/2021ImpactDelayedResponse/body.tex
fixing formatting

Accordingly, in the following we review background work in the areas of 
\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item[(\cref{paper:olguinmunoz2021impact:ssec:timeperception})] time perception
  \item[(\cref{paper:olguinmunoz2021impact:ssec:potentialmechs})] mechanisms relating human performance to delay.
\end{enumerate*}

@@

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
    \item arousal can expand durations~\cite{droit_volet2011emotion}.
\end{enumerate*}

@@

We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{chen2015early}.
This application belongs to a category of \gls{WCA}s designed to guide users through the execution of a sequential task.
% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
% The results of the user performing these instructions are provided as inputs to the system.
% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.

@@

The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
In particular
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
    \item it is easily understood by users and requires essentially no training
    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
    \item each step has an intrinsic hierarchical cognitive structure (see \cref{paper:olguinmunoz2021impact:fig:lego:hierarchical}), affording multiple levels of cognitive control.
\end{enumerate*}

@@

\subsubsection{Biometric Data}
The participants wore devices to acquire four physiological measures:
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item \glsreset{GSR}\gls{GSR}
  \item accelerometer data from the dominant wrist
  \item brain activity in the form of \glsreset{EEG}\gls{EEG}
  \item heart rate.
\end{enumerate*}

@@

For the \gls{EEG} data we employed the OpenBCI \gls{EEG} Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
It provides a quick and non-invasive way of obtaining \gls{EEG} signals from participants.
Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
The signals were sampled at \SI{200}{\hertz} and postprocessed to 
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
  \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
\end{enumerate*}

@@

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{enumerate*}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{paper:olguinmunoz2021impact:fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

@@

\subsection{\glsfmtshort{EEG}}
Analyses were conducted on the log \gls{EEG} power in the alpha band, beta band, and total of all bands measured.
Readings from the two frontally placed poles were highly correlated and were pooled into an average.
Logs were taken for analysis because the \gls{EEG} distribution tended to have a rightward tail.
Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 standard deviations from the mean of all participants).

@@

\subsection{\glsfmtshort{GSR} and Heart Rate}
The measure of \gls{GSR} was the log of total amplitude in the signal.
To control for effects of block length (steps~\( \times \)~delay) and the slow-down in execution time at longer delays, \gls{GSR} data were normalized by the temporal duration of the block.
Due to sensor failure and the elimination of one subject with extreme values of \gls{GSR} amplitude using the same rule as for \gls{EEG}, the analysis comprised 34 participants.
No systematic trend due to delay or block length was observed.
In addition, the same analysis of slice position within block length as was performed for the \gls{EEG} revealed no effects, indicating that \gls{GSR} was stable across step positions within a block.

>>>>

Accordingly, in the following we review background work in the areas of 
\begin{inlineenum}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item[(\cref{paper:olguinmunoz2021impact:ssec:timeperception})] time perception
  \item[(\cref{paper:olguinmunoz2021impact:ssec:potentialmechs})] mechanisms relating human performance to delay.
\end{inlineenum}

@@

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
    \item arousal can expand durations~\cite{droit_volet2011emotion}.
\end{inlineenum}

@@

We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{chen2015early}.
This application belongs to a category of \gls{WCA}'s designed to guide users through the execution of a sequential task.
% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
% The results of the user performing these instructions are provided as inputs to the system.
% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.

@@

The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
In particular
\begin{inlineenum}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
    \item it is easily understood by users and requires essentially no training
    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
    \item each step has an intrinsic hierarchical cognitive structure (see \cref{paper:olguinmunoz2021impact:fig:lego:hierarchical}), affording multiple levels of cognitive control.
\end{inlineenum}

@@

\subsubsection{Biometric Data}
The participants wore devices to acquire four physiological measures:
\begin{inlineenum}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item \glsreset{GSR}\gls{GSR}
  \item accelerometer data from the dominant wrist
  \item brain activity in the form of \glsreset{EEG}\gls{EEG}
  \item heart rate.
\end{inlineenum}

@@

For the \gls{EEG} data we employed the OpenBCI \gls{EEG} Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
It provides a quick and non-invasive way of obtaining \gls{EEG} signals from participants.
Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
The signals were sampled at \SI{200}{\hertz} and postprocessed to 
\begin{inlineenum}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
  \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
\end{inlineenum}

@@

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{inlineenum}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{inlineenum}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{paper:olguinmunoz2021impact:fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

@@

\subsection{\acs{EEG}}
Analyses were conducted on the log \gls{EEG} power in the alpha band, beta band, and total of all bands measured.
Readings from the two frontally placed poles were highly correlated and were pooled into an average.
Logs were taken for analysis because the \gls{EEG} distribution tended to have a rightward tail.
Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 standard deviations from the mean of all participants).

@@

\subsection{\acs{GSR} and Heart Rate}
The measure of \gls{GSR} was the log of total amplitude in the signal.
To control for effects of block length (steps~\( \times \)~delay) and the slow-down in execution time at longer delays, \gls{GSR} data were normalized by the temporal duration of the block.
Due to sensor failure and the elimination of one subject with extreme values of \gls{GSR} amplitude using the same rule as for \gls{EEG}, the analysis comprised 34 participants.
No systematic trend due to delay or block length was observed.
In addition, the same analysis of slice position within block length as was performed for the \gls{EEG} revealed no effects, indicating that \gls{GSR} was stable across step positions within a block.
